---
title: "MFRE Data Analytics Workshop Series"
subtitle: "Workshop 4: Python II"
author: "Krisha Lim"
date: "University of British Columbia | Master of Food and Resource Economics November 15, 2021" #"`r format(Sys.time(), '%d %B %Y')`"
output:
  xaringan::moon_reader:
    css: [default, metropolis, metropolis-fonts] 
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      highlightSpans: false
      countIncrementalSlides: false
---

```{css, echo=FALSE}
# CSS for including pauses in printed PDF output (see bottom of lecture)
@media print {
  .has-continuation {
    display: block !important;
  }
}

.gray { color: gray;}
```

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
library(knitr)
opts_chunk$set(
  fig.align="center",  
  fig.height=4, #fig.width=6,
  # out.width="748px", #out.length="520.75px",
  dpi=300, #fig.path='Figs/',
  cache=T#, echo=F, warning=F, message=F
  )

pacman::p_load(tidyverse, dplyr, reticulate)
reticulate::repl_python()
# reticulate::install_miniconda()
```

```{r, echo = F}
# IMPORTANT TO DO THIS FIRST
Sys.setenv(RETICULATE_PYTHON = "C:/Users/krishlim/AppData/Local/Programs/Python/Python310") 
# indicate that we want to use a specific condaenv
library(reticulate)
# py_config() 
#use_condaenv("r-reticulate")

# import SciPy (will use "r-reticulate" as per call to use_condaenv)
pandas <- import("pandas")
```

# Workshop preparation

☑ Make sure you have shared with me your completed `Workshop3_Student.ipynb` in Google Colab

☑ Review Workshop 3 - Python I notes and codes

☑ Upload and open the `Workshop4_Student.ipynb` file in Google Colab

---

# Overview

.gray[
**Python I**

  - Google Colab
  - Variable Types - String, Integer, Float, Boolean
  - Lists, Dictionaries, Functions

]

**Python II**

  - Importing libraries in Python
  - Importing files to Google Colab 
  - Working with data and Python packages

---

# Learning Outcomes

- Formulate simple function  ** only if not covered in Workshop 1

- Load data into Colab from a variety of sources

- Perform basic data cleaning and transformation steps

- Leverage Python library functions

---
class: inverse, center, middle
name: get_started

# Python I

<html><div style='float:left'></div><hr color='#EB811B' size=1px width=796px></html>

---
# Review - Lists

- Use **square brackets** to define lists
- Start at **index = 0**
- Use `list_name[start_included:end_excluded]` to access lists

```{python}
carbon = [15.3, 16.6, 7.06, 9.14]

# extract second and third values
print(carbon[1:3])
```
---
# Review - Dictionaries

- Use **curly brackets** to define dictionaries
- Contain *key:value* pairs

```{python}
data = {
  'canada': {'gdp':44100,
            'carbon':15.3},
  'usa': {'gdp':55700,
          'carbon':16.6}
}

# print canada's carbon emissions
print("Canada's carbon emissions = ", data['canada']['carbon'])
```
---
class: inverse, center, middle
name: get_started

# Python II

<html><div style='float:left'></div><hr color='#EB811B' size=1px width=796px></html>

---
# Data

- 3 months of historical sales of a supermarket with 3 branches in Myanmar (formerly Burma) 
- Downloaded from [Kaggle](https://www.kaggle.com/)
- **Add more here**

---
# About Packages

A library in Python contains a set of tools (called functions) that perform tasks on our data. 

Importing a library is like getting a piece of lab equipment out of a storage locker and setting it up on the bench for use in a project. 

Once a library is set up, it can be used or called to perform the task(s) it was built to do.

---
# About Packages

**Modules**
- collection of related code to keep codes organized

**Packages**
- collection of modules
- Examples: 
  - [`NumPy`](https://numpy.org/) - scientific computing
  - [`pandas`](https://pandas.pydata.org/) - processing data

**Libraries**
- usually refers to a collection of packages (or related modules and packages)
- Examples: 
  - [`matplotlib`](https://matplotlib.org/) - data visualization
  - [`TensorFlow`](https://www.tensorflow.org/resources/libraries-extensions) - machine learning

.footnote[
[1] Kateryna Koidan, ["Difference between Python modules, packages, librariers, and frameworks"](https://learnpython.com/blog/python-modules-packages-libraries-frameworks/)
]

---
# Pandas

- The [Python Data Analysis Library](https://pandas.pydata.org/) (a.k.a. Pandas) is one of the best and most famous packages working with tabular data in Python. The Pandas library provides data structures, produces high quality plots with [`matplotlib`](https://matplotlib.org/) and integrates nicely with other libraries that use [`NumPy`](https://numpy.org/) arrays. 

---
# Installing packages

- `pip` is Python's package manager
  - Need to frontend with `!` to execute a shell/cmd command in Colab
  
- `!pip list` prints out the lists of Python packages that is already in Colab

- `!pip install [package_name]` to install packages / libraries
  - `!pip install pandas` or `!pip install matplotlib`
  - `!pip install pandas matplotlib` to install multiple packages at once

---
# Importing packages

- In Python, we use the `import` function to import packages, modules, and libraries
  - It's like we have to use `library()` to load a package we have already installed using `install.packages()` in R
  
- Some syntax
  - `import numpy` -> Import the package Numpy and its functions
  - `import numpy as np` -> Shortens the imported name to `np` for better code readability
  - `from numpy import arange` -> Imports the `arange` function from `NumPy`

.footnote[
[1] [NumPy: the absolute basics for beginners](https://numpy.org/doc/stable/user/absolute_beginners.html)
]

---
# Using functions in packages

- Each time we call a function, we use the syntax `LibraryName.FunctionName`. 

- Adding the library name with a `.` before the function name tells Python where to find the function. 

---
# Importing modules or packages

```{python}
import numpy
a = numpy.arange(6)
a
```

```{python}
import numpy as np
a = np.arange(6)
a
```

```{python}
from numpy import arange
a = arange(6)
a
```

---
# Importing modules or packages

When to use which syntax? *It depends*

- If you will use only one function (e.g., `arange`) from the package multiple times, then use `from numpy import range`.

- If you will use multiple functions from the package, then use `import numpy as np`.

- There is actually another way (`from numpy import *`) but is not advisable when there are overlapping function names from different modules or packages

.footnote[
[1] Read [here](https://www.quora.com/What-is-the-difference-between-import-numpy-and-from-numpy-import-*) for more details
]

---
# Data Import - Local Drive

- Say you have the `.csv` or `.xlsx` file stored in your device

- Run this code first

```{python, eval = F}
# import pandas
import pandas as pd

# initiate file import 
from google.colab import files
uploaded = files.upload()
```

- A pop up will appear where you can select the file you want to work with. 

- Use `df.read_csv()` to read `.csv` files and `df.read_xlsx()` to read `.xlsx` files as `pandas` dataframe

```{python, eval = F}
# load file to object called 'supermarket'
supermarket = pd.read_csv("myanmar-supermarket.csv")
supermarket.head(5)
```

---
# Data Import - Google Drive

- Say you have the `.csv` or `.xlsx` file stored your Google Drive

- Run this code to mount your Google Drive

```{python, eval = F}
from google.colab import drive
drive.mount('/content/drive')
```

- Click the link, give authorization, and copy the authorization code and paste into the textbox in Colab

```{python, eval = F}
# load file to object called 'from_drive'
from_drive = pd.read_csv("/content/drive/MyDrive/myanmar-supermarket.csv")
from_drive.head(5)
```

---

# Data Import - Google Drive

- Say you want to import a lot of files from the same directory in Google Drive

- It would be better to specify the working directory to your Google Drive folder
  - We will use a package called `os` which will allow us to get the working directory

```{python, eval = F}
# import os package
import os

# print the current wd
os.getcwd()

# We want the dir to be "/content/drive/MyDrive" 

# if current wd is "/content" then 
dir = os.getcwd() + "/drive/MyDrive/" 

# if current wd is "/content/drive/" 
dir = os.getcwd() + "/MyDrive/"
 
# load file using `dir` in file path to object called `from_drive_path`
from_drive_path = pd.read_csv(dir + "myanmar-supermarket.csv")
from_drive_path.head(5)
```

---
# Data Import - Package/API

- Recall we used the [{`cansim`}](https://mountainmath.github.io/cansim/) package to access Statistics Canada data in R during the R bootcamp and Workshop 2 (Data Visualization)

- The equivalent in Python is [`stats_can`](https://stats-can.readthedocs.io/en/latest/)

```{python, eval = F}
# install package
!pip install stats-can

# change the directory to MyDrive
%pwd # prints the working directory
%cd /content/drive/MyDrive/ # set wd to MyDrive (if you haven't done earlier)

# import stats_can package from StatsCan 
from stats_can import StatsCan

# instantiate a StatsCan object
sc = StatsCan()

# read statscan table into a panda dataframe
df = sc.table_to_df("36-10-0222-01")
df.head(5)
```

---
# Dataframes

- A Dataframe is a 2-dimensional data structure that can store different types of data, such as characters, integers, floats, factors, etc. It is similar to an Excel spreadsheet or an SQL table or the `data.frame` in R. 

- As in Python lists, a dataframe always has an index that starts at 0. 

- An index refers to the position of an element in the data structure.

---
# Dataframes

```{python, echo = F}
import pandas as pd
supermarket = pd.read_csv("C:/Users/krishlim/OneDrive - UBC/Desktop/mfre/2021w-python-bootcamp/data/myanmar-supermarket.csv")
```

```{python}
supermarket.head(5)
```

---
# Exploring our data

- Here are a few useful functions to view dataframe objects in Python

```{python, eval = F}
# check type of thing `supermarket` object is
type(supermarket)

# what kind of things does `supermarket` contain? 
supermarket.dtypes

# access column names of `supermarket`
supermarket.columns

# access the shape/dimension of `supermarket`
supermarket.shape

# get first 15 obs of `supermarket`
supermarket.head(15)

# get last 15 obs of `supermarket`
supermarket.tail(15)
```

---
# Calculating summary statistics


---
# Recap

- Formulate simple function  ** only if not covered in Workshop 1

- Load data into Colab from a variety of sources

- Perform basic data cleaning and transformation steps

- Leverage Python library functions

---
# What's next?

- UBC MDS' [Programming in Python for Data Science](https://prog-learn.mds.ubc.ca/en/module0/)

---
# References

- [Data Analysis and Visualization in Python for Ecologists](https://datacarpentry.org/python-ecology-lesson/00-before-we-start/index.html) by Data Carpentry




```{r convert, include = F}
# library(xaringanBuilder)
# 
# build_pdf("C:/Users/krishlim/OneDrive - UBC/Desktop/mfre/2021w-python-bootcamp/presentation/workshop3.html")
# pagedown::chrome_print("C:/Users/krishlim/OneDrive - UBC/Desktop/mfre/2021w-python-bootcamp/presentation/workshop3.Rmd")
```
